Introduction to Markov state modeling with the PyEMMA software --- v\versionnumber

### 2018-07-25 manuscript.tex

In this tutorial, we assume the reader's familiarity with the basic theory behind the MSM approach (see Sec.~\ref{sec:background}) and focus on usage of PyEMMA.
Nevertheless, we will mention important theoretical concepts when appropriate throughout the tutorial.
We also assume the reader is familiar with molecular dynamics analyses of proteins and peptides and commonly used structural components of these systems; for a review of the simulation of biomolecular macromolecules, see~\cite{dror2012biomolecular}.

@@

In order to guide the user through the process before going into technical details,
the first notebook goes through the full analysis pipeline for,
we analyze the backbone conformation dynamics of a pentapeptide (Trp-Leu-Ala-Leu-Leu, Fig.~\ref{fig:io-to-ck}a) using 25 independently simulated trajectories with implicit solvent and a trajectory saving interval of $0.1$ ns~\cite{pyemma}.

\subsection{Feature selection}

@@

To begin, we start with the raw trajectory data, and we seek to construct and validate an MSM, which is represented by a transition matrix.

It is initially unknown which features (e.g. distances, contact maps, angles, etc.) are good descriptors of the slow CVs, i.e., are best suited to resolve the long-time dynamics~\cite{NoeClementiReview}.
A systematic way of making this choice has been introduced through the variational principle of conformation dynamics~\cite{noe-vac,nueske-vamk} and recently the more general version of non-equilibrium data, VAMP~\cite{vamp-preprint}.
VAMP provides a score that can be calculated for each feature in order to find good indicators of the long-time dynamics.
This score is computed via cross-validation in order to balance underfitting (too few features) and overfitting (too many features)~\cite{gmrq}.
Here we utilize the VAMP-2 score, which maximizes the kinetic variance of the estimated dynamical operator~\cite{kinetic-maps}.
To choose among three different molecular features relevant to protein structure, we compute the (cross-validated) VAMP-2 score at lag time $0.5$~ns and find that backbone torsions contain more kinetic variance than the backbone's heavy atom positions or distances between them (Fig.~\ref{fig:io-to-ck}b).
We note that recently, deep learning approaches for feature selection have been developed that may eventually replace the feature selection step~\cite{vampnet,tae,hernandez-vde}.

@@

\subsection{Discretization}

We then proceed with a $k$-means-based discretization of the four dimensional TICA subspace using $k=75$ cluster centers and perform the first validation test, i.e., the estimation of implied timescales (ITS) $t_i$, $i=0, 1,\dots$~\cite{swope-its}
(we note that the number of cluster centers can additionally be chosen using the VAMP score).
The ITS $t_i$ approximates the decorrelation time of the $i^\textrm{th}$ process and is computed from the eigenvalues $\lambda_i$ of the MSM transition matrix via
\begin{equation}
\label{eq:its}
t_i = \frac{-\tau}{\ln\left|\lambda_i(\tau)\right|}
\end{equation}
with $\tau$ being the lag time.
In this validation test, we look for ITS convergence and choose $\tau$ accordingly, i.e., within a range where the ITS are approximately invariant; the uncertainty of the implied timescales is quantified based upon Markov models sampled according to a Bayesian scheme~\cite{ben-rev-msm,noe-tmat-sampling} with $100$ samples.
In our example, we find the three slowest ITS to converge quickly and to be constant within a $95\%$ confidence interval for lag times above $0.5$ ns (Fig.~\ref{fig:io-to-ck}e).
We can further see a converged fourth process (cyan line); however, this process decays for lag times above $2.5$ ns and thus might not be well resolved in the final model.

\subsection{MSM estimation and validation}

Finally, we estimate a (Bayesian) MSM at a the lag time $\tau=0.5$ ns and validate the model using a Chapman-Kolmogorov (CK) test, which compares the right and the left side of the Chapman-Kolmogorov equation
\begin{equation}
\label{eq:ck}
T(k \tau) = T^k(\tau)
\end{equation}
with $T$ being the MSM transition matrix.
To perform this test, we first choose a number of metastable states appropriate for our system. Then, PyEMMA automatically estimates additional MSM transition matrices at lag times $k \tau$ and compares these estimates to the $k^\textrm{th}$ power of the original transition matrix $T(\tau)$.
As we have identified three slow relaxation timescales in the ITS test,
we perform the CK test on the four metastable states whose kinetics involves these three relaxation timescales.
The CK test shows indeed that our model estimated at lag time tau predicts the kinetic behavior on longer timescales accurately (Fig.~\ref{fig:io-to-ck}f).
We note that our MSM estimator results in a model that adheres to detailed balance.

@@

A fuzzy spectral clustering using the PCCA++ algorithm~\cite{pcca++} allows us to coarse grain the $k$-means microstates into four metastable macrostates (Fig.~\ref{fig:msm-analysis}b) $\mathcal{S}_i$, $i=1,\dots,4$, for which we then approximate the stationary probabilities and relative free energies (defined up to an additive constant)
\[ \begin{array}{ccc}
\textrm{macrostate } \mathcal{S}_i & \pi_{\mathcal{S}_i} & G_{\mathcal{S}_i} / \textrm{k}_\textrm{B} T \\
\hline
\mathcal{S}_1 & 0.004 & 5.565 \\
\mathcal{S}_2 & 0.013 & 4.314 \\
\mathcal{S}_3 & 0.021 & 3.864 \\
\mathcal{S}_4 & 0.962 & 0.039
\end{array}\]
using the relation
\begin{equation}
\label{eq:fe}
G_{\mathcal{S}_i} = - \textrm{k}_\textrm{B} T \ln \sum\limits_{j\in \mathcal{S}_i} \pi_j,
\end{equation}
where $\pi_j$ denotes the MSM stationary weight of the $j^\textrm{th}$ microstate.

@@

\subsection{Connecting the MSM with experimental data}

To illustrate how to connect an MSM analysis with experiments, we compute the Trp-1 solvent accessible surface area (SASA) using the mdtraj~\cite{mdtraj} package.
The SASA is a reasonable approximation for the instantaneous tryptophan florescence signal.
The auto correlation function (ACF) of this signal can be measured in spectroscopy experiments.

In the projection onto the first two TICA components, we observe a high variance of the SASA in all macrostates except $\mathcal{S}_1$ (Fig.~\ref{fig:msm-analysis}e).
This indicates that Trp-1 has only a low solvent exposure in this macrostate.
The ACF itself yields a weak signal (Fig.~\ref{fig:msm-analysis}f, top panel).
If, however, we prepare the system in macrostate $\mathcal{S}_1$ (non-equilibrium condition), the relaxation towards equilibrium yields a much stronger signal (Fig.~\ref{fig:msm-analysis}f, bottom panel).

Finally, we sample $200$ super-imposed molecular conformations from each of the $4$ macrostates (Fig.~\ref{fig:msm-analysis}g) and render them using VMD~\cite{vmd}.
We observe that, in macrostates $\mathcal{S}_1$ and $\mathcal{S}_2$, the backbone appears to often be in close proximity to the Trp-1 (shown in red), thus effectively limiting Trp-1 exposure to the solvent.
In the (more probable) macrostates $\mathcal{S}_3$ and $\mathcal{S}_4$, the Trp-1 appears to be much more exposed.

@@

In this section, we have summarized how to conduct an MSM-based analysis of biomolecular dynamics data using PyEMMA. For the full analysis, please refer to the first notebook (00). All notebooks as well as detailed installation instructions are available on \githubrepository{}.

\section{Author Contributions}
%%%%%%%%%%%%%%%%
% This section mustt describe the actual contributions of
% author. Since this is an electronic-only journal, there is
% no length limit when you describe the authors' contributions,
% so we recommend describing what they actually did rather than
% simply categorizing them in a small number of
% predefined roles as might be done in other journals.
%
% See the policies ``Policies on Authorship'' section of https://livecoms.github.io
% for more information on deciding on authorship and author order.
%%%%%%%%%%%%%%%%
CW, MKS, TH, SO, and FN designed research.
CW, MKS, TH, BEH, and SO developed and tested notebooks.
MKS developed the software infrastructure, test, and install environment.
CW, MKS, TH, BEH, and FN wrote the manuscript.

>>>>

In this tutorial, we assume the reader's familiarity with the basic theory behind the MSM approach (see Sec.~\ref{sec:background}) and focus on usage of PyEMMA.
Nevertheless, we will mention important theoretical concepts when appropriate throughout the tutorial.
We also assume the reader is familiar with molecular dynamics analyzes of proteins and peptides and commonly used structural components of these systems; for a review of the simulation of biomolecular macromolecules, see~\cite{dror2012biomolecular}.

@@

For the remainder of this manuscript we will walk through the first notebook (00). In notebook 00 we analyze a dataset of the Trp-Leu-Ala-Leu-Leu pentapeptide, consisting of 25 independent MD trajectories conducted in implicit solvent with a saving interval of 0.1~ns. We present the results obtained in the notebook, thereby providing an example of how results generated using PyEMMA can be integrated into research publications.
\subsection{Feature selection}

@@

In Markov state modeling our objective is to model the slow dynamics of a molecular process. In order to approximate the slow dynamics in a statistically efficient manner, a lower dimensional representation of our simulation data is necessary. However, the features (e.g. torsion angles, distances or contacts) which best represent the slow dynamical modes of a given molecular system is unknown a priori~\cite{NoeClementiReview}. Fortunately, the variational principle of conformational dynamics~\cite{noe-vac,nueske-vamk} and the more general variational approach for Markov processes (VAMP)~\cite{vamp-preprint} provides a systematic means to quantitatively compare multiple representations of the dataset. In particular, we can use the VAMP score to directly compare the ability of certain features to capture slow dynamical modes in a particular molecular system.
%Here, we use the VAMP score in a cross-validated manner to ensure we neither include too few features (under-fitting) or too many features (over-fitting)~\cite{gmrq}.

Here we utilize the VAMP-2 score, which maximizes the kinetic variance of the estimated dynamical operator~\cite{kinetic-maps}.
We should always evaluate the score in a cross-validated manner to ensure we neither include too few features (under-fitting) or too many features (over-fitting)~\cite{gmrq,vamp-preprint}.
To choose among three different molecular features relevant to protein structure, we compute the (cross-validated) VAMP-2 score at lag time $0.5$~ns and find that backbone torsions contain more kinetic variance than the backbone's heavy atom positions or distances between them (Fig.~\ref{fig:io-to-ck}b).
We note that recently, deep learning approaches for feature selection have been developed that may eventually replace the feature selection step~\cite{vampnet,tae,hernandez-vde}.

@@

\subsection{Discretization}
TICA yields a representation of our molecular simulation data with a reduced dimensionality, which can greatly facilitate the decomposition of our system into the discrete Markovian states necessary for MSM estimation. Here, we use the $k$-means algorithm to segment the four dimensional TICA into $k=75$ cluster centers. The number of cluster centers has been chosen to optimize the VAMP-2 score as in a manner identical to how the feature selection was carried out above, but is not presented in the showcase notebook.

\subsection{MSM estimation and validation}
When estimating a MSM it is critical to chose a lag time, $\tau$, which is long enough to ensure Markovian dynamics in our reduced space. Simultaneously we want to keep $\tau$ short to conserve as much of our MD data as possible. Plotting the implied timescales (ITS) as a function $\tau$ can be a helpful diagnostic when selecting the MSM lag time~\cite{swope-its}. The ITS $t_i$ approximates the decorrelation time of the $i^\textrm{th}$ process and is computed from the eigenvalues $\lambda_i$ of the MSM transition matrix via
\begin{equation}
\label{eq:its}
t_i = \frac{-\tau}{\ln\left|\lambda_i(\tau)\right|}
\end{equation}
with $\tau$ being the lag time.
A necessary condition for Markovian dynamics in our reduced space is that the ITS are approximately constant as a function of $\tau$; accordingly, we chose the smallest possible $\tau$ which fulfills this condition within the model uncertainty. The uncertainty bounds are computed using a Bayesian scheme~\cite{ben-rev-msm,noe-tmat-sampling} with $100$ samples.
In our example, we find that the four slowest ITS converge quickly and are constant within a $95\%$ confidence interval for lag times above $0.5$~ns (Fig.~\ref{fig:io-to-ck}e). Using this lag time we can now estimate a (Bayesian) MSM with $\tau=0.5\mathrm{ns}$.

To test the validity of our MSM we perform a Chapman-Kolmogorov (CK) test. The CK test compares the right and the left side of the Chapman-Kolmogorov equation
\begin{equation}
\label{eq:ck}
T(k \tau) = T^k(\tau)
\end{equation}
with $T$ being the MSM transition matrix. The left-hand side of the equation corresponds to a MSM estimated at lag time $k\tau$ with $k$ being an integer larger than 1, whereas the right-hand side of the equation is our estimated MSM to the $k^\textrm{th}$ power. Visualizing the full transition probability matrix $T$ is difficult; we therefore coarse-grain $T$ into a number of metastable states before performing the test. An appropriate number of metastable states can be chosen by identifying a relatively large gap in the ITS plot. In our case we have multiple gaps and therefore various options choose from. We chose 5 metastable states for this analysis, as it is the largest number that only includes processes which are well above the MSM lag time.
Given the number of metastable states, PyEMMA automatically estimates additional MSM transition matrices at lag times $k \tau$ and compares these estimates to the $k^\textrm{th}$ power of the original transition matrix $T(\tau)$ using a coarse-grained representation.
The CK test shows that the MSM we have estimated at lag time $\tau=0.5$~ns indeed predicts the long-timescale behavior of our system (Fig.~\ref{fig:io-to-ck}f).

@@

A spectral clustering using the PCCA++ algorithm~\cite{pcca++} allows us to coarse-grain the $k$-means microstates into five metastable macrostates (Fig.~\ref{fig:msm-analysis}b) $\mathcal{S}_i$, $i=1,\dots,5$, for which we then approximate the stationary probabilities and relative free energies (defined up to an additive constant)
\[ \begin{array}{ccc}
\textrm{macrostate } \mathcal{S}_i & \pi_{\mathcal{S}_i} & G_{\mathcal{S}_i} / \textrm{k}_\textrm{B} T \\
\hline
\mathcal{S}_1 & 0.004 & 5.565 \\
\mathcal{S}_2 & 0.013 & 4.314 \\
\mathcal{S}_3 & 0.021 & 3.864 \\
\mathcal{S}_4 & 0.962 & 0.039 \\
\mathcal{S}_5 & \text{TODO} & \text{TODO} \\
\end{array}\]
using the relation
\begin{equation}
\label{eq:fe}
G_{\mathcal{S}_i} = - \textrm{k}_\textrm{B} T \ln \sum\limits_{j\in \mathcal{S}_i} \pi_j,
\end{equation}
where $\pi_j$ denotes the MSM stationary weight of the $j^\textrm{th}$ microstate.

@@

\subsection{Connecting the MSM with experimental data}
MSMs can also be connected to experimental observables. Connecting MSM analysis to experimental data can both serve as an accuracy test of our MSM as well as provide a mechanistic interpretation of observed experimental signals.
Since we have both the stationary and dynamic properties of the molecular system encoded into the transition probability matrix, we can compute observables that involve both stationary ensemble averages as well as correlation functions.

As an example, here we look at the fluorescence correlation of the tryptophan in our system, Trp-1, since this is a realistic experimental observable for our pentapeptide system.
In order to compute the fluorescence correlation functions we require a microscopic, instantaneous value of the tryptophan fluorescence for each Markov state.
%These values are computed using a 'forward model'.
To approximate the fluorescence signal in our pentapeptide system, we use the mdtraj library~\cite{mdtraj} to compute the solvent accessible surface area (SASA) of Trp-1.

Now that we have an approximation of the fluorescence in each of our Markovian states, we can use PyEMMA to compute the fluorescence auto-correlation function (ACF) from our MSM (\ref{fig:msm-analysis}e, upper).
Note how the computed ACF has a very small response (i.e., signal amplitude).
Using PyEMMA, we can simulate the relaxation of an observable if we had we prepared our molecular system in a non-equilibrium initial condition.
The experimental counterpart of such prediction could be temperature or pressure jump experiments or a stopped flow assay.
To illustrate such an experiment, we initialize our molecular ensemble as the metastable distribution of $\mathcal{S}_1$ and follow the predicted fluorescence signal as it relaxes to equilibrium (\ref{fig:msm-analysis}e, lower).
We see that the predicted relaxation signal has a much larger amplitude for the non-equilibrium initialization, making it more likely to be experimentally measurable.

@@

In this section, we have summarized how to conduct a MSM-based analysis of biomolecular dynamics data using PyEMMA. For the full analysis, please refer to the first notebook (00). All notebooks as well as detailed installation instructions are available on \githubrepository{}.

\section{Author Contributions}
%%%%%%%%%%%%%%%%
% This section must describe the actual contributions of
% author. Since this is an electronic-only journal, there is
% no length limit when you describe the authors' contributions,
% so we recommend describing what they actually did rather than
% simply categorizing them in a small number of
% predefined roles as might be done in other journals.
%
% See the policies ``Policies on Authorship'' section of https://livecoms.github.io
% for more information on deciding on authorship and author order.
%%%%%%%%%%%%%%%%
CW, MKS, TH, SO, and FN designed research.
CW, MKS, TH, BEH, and SO developed and tested notebooks.
MKS developed the software infrastructure, test, and install environment.
CW, MKS, TH, BEH, SO, and FN wrote the manuscript.
